---
title: "CDISC SDTMIG Conformance Validation"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(pointblank)
library(dplyr)
```

The **pointblank** package includes a built-in CDISC conformance engine that can
validate SDTM datasets against the SDTMIG rule catalog. This engine runs
entirely within R, so no external tools or API calls are needed at runtime. The
CDISC CORE rules and SDTM Controlled Terminology are bundled as JSON files
inside the package, so validation works offline and out of the box.

The conformance engine evaluates 426 rules from the SDTMIG v3.4 specification.
These cover per-row value checks (e.g., codelist membership, ISO 8601 date
formats), structural checks (required variables, column ordering), and
domain-level requirements (required domains like DM, TS, and TA). Each rule
traces back to an official CDISC CORE rule identifier, making it straightforward
to cross-reference findings with published CDISC guidance.

## Quick Start

The main entry point is `validate_sdtmig()`. It takes a named list of data
frames (one per SDTM domain) and returns a `cdisc_conformance_result` object.
The names in the list correspond to domain abbreviations (e.g., `"DM"`, `"AE"`,
`"LB"`) and are matched case-insensitively.

Let's set up a small Demographics (`DM`) domain with a couple of deliberate
problems. The third subject has an invalid `SEX` value (`"INVALID"` is not in
the CDISC controlled terminology for SEX), and the second subject has a
malformed reference start date (`"not-a-date"` is not a valid ISO 8601 datetime
string).

```{r}
dm <- 
  dplyr::tibble(
    STUDYID  = rep("STUDY01", 3),
    DOMAIN   = rep("DM", 3),
    USUBJID  = c("STUDY01-001", "STUDY01-002", "STUDY01-003"),
    SUBJID   = c("001", "002", "003"),
    SEX      = c("M", "F", "INVALID"),
    AGE      = c(45, 62, 38),
    RFSTDTC  = c("2024-01-15", "not-a-date", "2024-03"),
    ARMCD    = c("TRT", "PBO", "TRT"),
    ARM      = c("Treatment", "Placebo", "Treatment"),
    ACTARMCD = c("TRT", "PBO", "TRT"),
    ACTARM   = c("Treatment", "Placebo", "Treatment"),
    COUNTRY  = c("USA", "GBR", "CAN")
  )
```

Now run the validation. The engine loads the bundled rule catalog and controlled
terminology automatically:

```{r}
result <- validate_sdtmig(list(DM = dm))

result
```

The printed summary tells us how many of the 426 rules passed, failed, or were
not applicable (rules that require inputs we didn't provide, like Define-XML
metadata or additional domains). In this case the engine flagged several issues:
the two data-quality problems we planted, plus a few structural findings
(missing required variables and domains that a complete submission would
include).

## Inspecting Results

The conformance result object has several accessor functions that make it easy
to drill into the details using familiar tidyverse patterns.

### Overall status

The `cdisc_all_passed()` function gives a single logical value: did every rule
pass without findings?

```{r}
cdisc_all_passed(result)
```

For a more granular view, `cdisc_status_counts()` breaks down the rules by
status. The five possible statuses are `pass`, `fail`, `error` (rule encountered
an unexpected problem), `not_applicable` (required inputs missing), and
`not_supported` (rule type not yet implemented).

```{r}
cdisc_status_counts(result)
```

### Issues summary

The `cdisc_issues()` function returns a data frame containing only the rules
that found problems. Each row represents one failing rule with its ID, the
affected dataset, the issue count, severity, and message. This is the best
starting point for understanding what went wrong at a glance.

```{r}
cdisc_issues(result) |>
  dplyr::select(rule_id, dataset, n_issues, sensitivity, message)
```

Notice that the issues include both the data-quality problems we introduced
(SDTM-007 for the invalid SEX value, SDTM-011 for the malformed date) and
structural findings (SDTM-032 through SDTM-034 for missing DM variables,
SDTM-121 and SDTM-122 for the required TS and TA domains that a complete
submission would contain).

### Row-level findings

For rules that perform per-row checks (like `RECORD_CHECK`), the engine captures
detailed findings that pinpoint exactly which subject, row, and column triggered
each violation. The `cdisc_findings_df()` function flattens these into a tidy
data frame:

```{r}
cdisc_findings_df(result) |>
  dplyr::select(rule_id, dataset, row, usubjid, checked_column, checked_value)
```

The `row` column uses 1-based indexing (following R convention), so you can look
up the offending row directly with `dm[row, ]`. The `checked_column` and
`checked_value` fields tell you exactly which variable failed and what value it
contained, making it straightforward to trace the problem back to source data.

## Working with Multiple Domains

Real SDTM submissions contain many domains. Pass them all as entries in the
named list and the engine will evaluate each rule against its applicable
domain(s). Rules that check for domain presence or cross-domain consistency will
also run as appropriate.

Here we add a small Adverse Events (`AE`) domain alongside the DM:

```{r}
ae <- 
  dplyr::tibble(
    STUDYID = rep("STUDY01", 2),
    DOMAIN  = rep("AE", 2),
    USUBJID = c("STUDY01-001", "STUDY01-002"),
    AESEQ   = c(1, 1),
    AETERM  = c("Headache", "Nausea"),
    AEDECOD = c("Headache", "Nausea"),
    AESTDTC = c("2024-02-01", "2024-02-15"),
    AEENDTC = c("2024-02-03", "2024-02-17")
  )

result_multi <- validate_sdtmig(list(DM = dm, AE = ae))

result_multi
```

With the AE domain present, rules that apply specifically to adverse event data
now run as well. The total rule count stays at 426 (the full catalog is always
loaded), but more rules will produce `pass` or `fail` results rather than being
skipped for lack of input data.

We can quickly compare which rules failed across the two runs:

```{r}
cdisc_issues(result_multi) |>
  dplyr::select(rule_id, dataset, n_issues, message)
```

## Filtering by Rule Type

Sometimes you only want to focus on a particular category of checks. The
`rule_types` argument lets you restrict which rule types are loaded and
evaluated. This can be useful during iterative data cleaning, where you might
want to fix all per-row value issues before moving on to structural checks.

```{r}
result_records <- 
  validate_sdtmig(
    list(DM = dm),
    rule_types = "RECORD_CHECK"
  )

result_records
```

The engine supports seven rule types. The first five are fully implemented and
the last two are stubs that will become active when a Define-XML importer is
added to the R package:

| Rule type | What it checks |
|:---|:---|
| `RECORD_CHECK` | Per-row value validation (codelist membership, date formats, consistency) |
| `VARIABLE_METADATA_CHECK` | Column presence and ordering within a domain |
| `DATASET_METADATA_CHECK` | Dataset-level attributes (sort keys, required structure) |
| `DATASET_CONTENTS_CHECK` | Dataset-level value constraints evaluated row-by-row |
| `DOMAIN_PRESENCE_CHECK` | Whether required domains are present or prohibited ones are absent |
| `DEFINE_ITEM_METADATA_CHECK` | Variable declarations against Define-XML (stub) |
| `DEFINE_CODELIST_CHECK` | Codelist values against Define-XML declarations (stub) |

## What's in the Rule Catalog?

The bundled catalog for SDTMIG v3.4 contains 426 rules translated from the CDISC
CORE rule definitions. Each rule consists of:

- A **core_id** (e.g., `"SDTM-007"`) that links back to the official CDISC CORE
rule identifier, making it easy to cross-reference with published CDISC
documentation and the CORE Rules Engine output.
- A **rule_type** that determines how the rule is dispatched by the engine
(record check, metadata check, etc.).
- A set of **operations** that compute intermediate columns before evaluation.
These handle things like codelist lookups, ISO 8601 format validation, data type
verification, column presence checks, and variable ordering.
- A **conditions** tree (a nested structure of AND/OR/NOT boolean expressions)
that defines when a violation is flagged. Conditions reference the computed
columns added by operations.
- A **sensitivity** level (`Error`, `Warning`, or `Notice`) indicating the
severity of a finding per CDISC guidance.

## Controlled Terminology

CDISC publishes Controlled Terminology (CT) packages that define the permitted
values for coded variables like `SEX`, `RACE`, `COUNTRY`, and many others. The
engine ships with a bundled SDTM CT package (`sdtm-ct-2024-09-27`) that is
loaded automatically when you call `validate_sdtmig()`.

Codelist validation is case-insensitive: a value of `"m"` will match the
codelist term `"M"`. Missing values (`NA`) and empty strings are always treated
as valid for codelist purposes (null-handling is covered by separate not-null
rules).

If additional CT packages are bundled in the future (e.g., newer terminology
releases or ADaM CT), you can select them explicitly with the `ct_packages`
argument:

```{r eval=FALSE}
result <- 
  validate_sdtmig(
    list(DM = dm),
    ct_packages = "sdtm-ct-2024-09-27"
  )
```

## Current Limitations

There are a few areas where the R conformance engine differs from the full CDISC
CORE Rules Engine or has functionality that is not yet implemented:

- **Define-XML support**: The four Define-XML-aware operations
(`define_var_declared`, `define_required_check`, `define_codelist_check`,
`define_type_check`) are currently stubs that always pass. They will produce
real results once a Define-XML importer is added to the R package. In the
meantime, rules that depend on Define-XML metadata will simply not flag any
issues (they won't produce false positives).

- **SDTMIG version coverage**: Only SDTMIG v3.4 is currently bundled. Additional
standard/version combinations can be added by running the catalog generation
script from the Python **pointblank** package and copying the resulting JSON
file to the `inst/conformance/rules/` directory.

- **JSONata expressions**: A small number of advanced CORE rules use JSONata
expressions for complex cross-dataset logic. The JSONata evaluator is not yet
ported to R. Rules that depend on it will return `not_supported`. This affects a
very small fraction of the catalog.
