---
title: "YAML Validation Workflows"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(pointblank)
```

**pointblank** lets you define, store, and execute validation workflows entirely through YAML files. This article explains the YAML format and walks through the functions you use to read, write, and run YAML-based validations.

## Why use YAML?

Writing an agent and its validation plan as R code is the most direct approach, but YAML offers a distinct set of advantages:

- validation logic is stored in plain text that can be version-controlled alongside your data infrastructure
- the same YAML file can be reused across different R scripts, scheduled jobs, or CI pipelines
- non-programmers can read (and may prefer to edit) the YAML without needing to understand R
- YAML files are interoperable with **Python pointblank**, which accepts the same format

The YAML approach shouldn't replace the R API but, rather, it complements it. Most simple validations work well in YAML whereas complex logic with custom functions is much easier to express in R.

## The YAML file format

A **pointblank** agent YAML file has the following top-level structure:

```yaml
type: agent
tbl: ~ small_table                  # R expression for the target table
tbl_name: small_table               # Optional display name
label: "My validation plan"         # Optional label for the report
lang: en                            # Language for report text
locale: en                          # Locale for number formatting
actions:
  warn_fraction: 0.1                # Global action thresholds
  stop_fraction: 0.2
steps:
  - col_exists:                     # Validation function name (as key)
      columns: vars(date)
  - col_vals_gt:
      columns: vars(d)
      value: 100.0
  - col_vals_le:
      columns: vars(c)
      value: 5.0
  - col_vals_regex:
      columns: vars(b)
      pattern: '[0-9]-[a-z]{3}-[0-9]{3}'
```

The top-level fields map directly to arguments of `create_agent()`:

| YAML key | Corresponds to |
|---|---|
| `tbl` | `tbl` argument of `create_agent()` (an R expression) |
| `tbl_name` | `tbl_name` argument of `create_agent()` |
| `label` | `label` argument of `create_agent()` |
| `lang` | `lang` argument of `create_agent()` |
| `locale` | `locale` argument of `create_agent()` |
| `actions` | `actions` argument of `create_agent()`: use `warn_fraction`, `stop_fraction`, `notify_fraction`, or their absolute-count variants (`warn_count`, etc.) |

The `steps` list maps each entry to a validation function call. The key is the full function name (e.g., `col_vals_gt`, `col_exists`, `rows_distinct`) and the value is a mapping of the function's named arguments.

## Creating a YAML file from R

The easiest way to produce a valid **pointblank** YAML file is to build an agent in R and write it to disk with `yaml_write()`.

```r
agent <-
  create_agent(
    tbl = ~ small_table,
    tbl_name = "small_table",
    label = "A simple example with the `small_table`.",
    actions = action_levels(warn = 0.1, stop = 0.2)
  ) |>
  col_exists(columns = vars(date, date_time)) |>
  col_vals_gt(columns = vars(d), value = 100) |>
  col_vals_le(columns = vars(c), value = 5) |>
  col_vals_regex(columns = vars(b), pattern = "[0-9]-[a-z]{3}-[0-9]{3}")

yaml_write(agent, filename = "agent-small_table.yml")
```

The resulting YAML file faithfully captures the agent and its entire validation plan. Interrogation data is **not** written to YAML. The file always represents a plan ready for a new run.

## Reading a YAML file back into an agent

Use `yaml_read_agent()` to reconstruct an agent object from a YAML file:

```r
agent <- yaml_read_agent(filename = "agent-small_table.yml")

agent
```

This gives you an agent with a validation plan already in place but without interrogation data (the report will helpfully note that no interrogation has been performed yet). You have the option of adding further validation steps to this agent before interrogating.

To interrogate the table in a single step:

```r
agent <- yaml_agent_interrogate(filename = "agent-small_table.yml")

agent
```

`yaml_agent_interrogate()` reads the YAML, builds the agent, calls `interrogate()`, and returns the agent with full intel.

**pointblank** provides a ready-to-use example YAML file. You can locate it with the `system.file()` function:

```r
yml_file_path <-
  system.file(
    "yaml", "agent-small_table.yml",
    package = "pointblank"
  )

agent <- yaml_agent_interrogate(filename = yml_file_path)
```

## Inspecting the expressions a YAML file generates

Before running anything, you can preview the **pointblank** R expressions that a YAML file will produce using `yaml_agent_show_exprs()`:

```r
yaml_agent_show_exprs(filename = "agent-small_table.yml")
```

This prints the equivalent R code to the console, which is useful for understanding how YAML keys map to function calls, and for debugging unexpected behavior.

## Writing the agent YAML string directly

If you want the YAML representation of an existing agent as a character string (rather than writing to disk), use `yaml_agent_string()`:

```r
agent |> yaml_agent_string()
```

## Validation steps in YAML

Every **pointblank** validation function has a direct YAML equivalent. The function name (e.g., `col_vals_gt`) becomes the YAML step key, and the function's arguments become the step's sub-keys:

```yaml
steps:
  # col_vals_gt(columns = vars(a), value = 0, na_pass = FALSE)
  - col_vals_gt:
      columns: vars(a)
      value: 0.0
      na_pass: false

  # col_vals_between(columns = vars(d), left = 0, right = 10000, inclusive = c(TRUE, TRUE))
  - col_vals_between:
      columns: vars(d)
      left: 0.0
      right: 10000.0
      inclusive:
        - true
        - true

  # col_vals_in_set(columns = vars(f), set = c("low", "mid", "high"))
  - col_vals_in_set:
      columns: vars(f)
      set:
        - low
        - mid
        - high

  # col_vals_regex(columns = vars(b), pattern = "^[0-9]-[a-z]{3}-[0-9]{3}$")
  - col_vals_regex:
      columns: vars(b)
      pattern: '^[0-9]-[a-z]{3}-[0-9]{3}$'

  # rows_distinct()
  - rows_distinct
```

Arguments that use R expressions (like column selections via `vars()`) are stored as strings and evaluated when the YAML is read back.

Here is a live example: we build an agent, write it to a temporary YAML file, read it back, interrogate, and check that it passes:

```{r}
tmp_yaml <- tempfile(fileext = ".yml")

create_agent(
  tbl = ~ small_table,
  label = "YAML round-trip example"
) |>
  col_vals_gt(columns = vars(d), value = 100) |>
  col_vals_le(columns = vars(c), value = 5) |>
  col_vals_in_set(columns = vars(f), set = c("low", "mid", "high")) |>
  yaml_write(filename = tmp_yaml, quiet = TRUE)

agent <- yaml_agent_interrogate(filename = tmp_yaml)

all_passed(agent)
```

### Per-step options

Each step can carry its own `actions`, `label`, `brief`, `step_id`, and `active` fields, mirroring the corresponding arguments of the validation functions:

```yaml
steps:
  - col_vals_not_null:
      columns: vars(a, b)
      label: "Core fields must be complete"
      active: true
      actions:
        warn_count: 1
        stop_fraction: 0.1
```

## Table stores and `tbl_source()`

When the same target table is used across many agents (or informants), it's convenient to define the table-prep formula once and reference it by name. This is what the *table store* (`tbl_store`) is for.

### Creating a table store

```r
store <-
  tbl_store(
    small_table ~ pointblank::small_table,
    small_table_sqlite ~ {
      db <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
      dplyr::copy_to(db, pointblank::small_table, "small_table")
      dplyr::tbl(db, "small_table")
    }
  )

yaml_write(store, filename = "tbl_store.yml")
```

The resulting `tbl_store.yml` looks like:

```yaml
type: tbl_store
tbls:
  small_table: ~ pointblank::small_table
  small_table_sqlite: ~{...}
```

### Referencing tables with `tbl_source()`

In an agent YAML file, use `tbl_source()` to refer to a named entry in a `tbl_store.yml`:

```yaml
type: agent
tbl: ~ tbl_source("small_table", "tbl_store.yml")
tbl_name: small_table
steps:
  - col_vals_gt:
      columns: vars(d)
      value: 100.0
```

At interrogation time, `tbl_source()` reads the table-prep formula from `tbl_store.yml` and evaluates it to produce the target table. This keeps the agent YAML decoupled from the specific connection details.

## Batch execution with `yaml_exec()`

For scheduled or pipeline-based quality monitoring, you can set up a directory of YAML files and run them all at once:

```r
yaml_exec(path = "my_validations/")
```

A typical directory might contain:

```
my_validations/
  tbl_store.yml               # Table store (optional but recommended)
  agent-customers.yml         # Agent for the customers table
  agent-orders.yml            # Agent for the orders table
  informant-customers.yml     # Informant for the customers table
  output/                     # Where results are written (created automatically)
```

`yaml_exec()` interrogates each agent YAML file and incorporates each informant YAML file, then writes the resulting objects to the output directory as timestamped RDS files. These can be read back later with `x_read_disk()` or, for multiple agents together, with `read_disk_multiagent()`.

```r
# Read the most recently saved agent back
agent <- x_read_disk("my_validations/output/agent-customers-2025-01-15_09-00-00.rds")

# Read multiple agent snapshots into a multiagent
multiagent <- read_disk_multiagent(path = "my_validations/output/")
```

The multiagent collects results across time, enabling trend analysis of data quality over multiple runs.
